Calculating the Solvent Exposed Surface Area of Proteins with Intersecting Spheres

### 2020-06-08 Paper.tex
final changes?

Our protein can be represented as a set of atom centers $S$ in $\R^3$ with their corresponding balls of various radii. One aspect of the model is computing the power diagram of this protein. The power diagram is a generalized Voronoi diagram, where instead of the normal Euclidean metric, the definition for distance between an atom with center $z_i$ and radius $\varrho_i$ and a point $x$ is $\pi_i(x)=\|x-z_i\|^2-\varrho_i^2$. This becomes the default Voronoi diagram of $S$ if all the radii are equal. After this power diagram is generated we can associate each set of atoms $S_i$ with its corresponding power cell $P_i$ defined by
\begin{equation*}
P_i=\{x\in\R^3\mid\pi_i(x)\leq\pi_j(x),\forall j\}
\end{equation*}
Each $P_i$ is convex, and together they cover all $\R^3$ without overlap. The dual of this power diagram is a weighted Delaunay Triangulation of $S$ into a set of simplices, or triangles and tetrahedrons. We can compute the surface area of these balls using Delaunay triangulation, which will help us in the later sections. There are additional rules for constructing Delaunay triangles, but for the purposes of this paper, it is only necessary to know how they are constructed from the atoms.

@@

Consider a set of balls $B_i$ defined by a center $z_i$ and a fixed radius $\varrho_i$ to be our protein. To represent the entire state of the protein (with $n$ atoms) we use the $m=3n$ dimensional  vector $\textbf{z}\in\R^{3n}$ where $\lbrack{\bf z}_{3i+1}, {\bf z}_{3i+2}, {\bf z}_{3i+3}\rbrack^T=z_i$. Now we can define the weighted and unweighted surface areas $E,A:\R^{3m}\rightarrow\R$ as real multivariate functions from the state of the protein to a single value (the solvent exposed surface area). The weighted area will be described below. Here we will describe the gradients of these maps, written as $DE_{\bf z}$ and $DA_{\bf z}$ which would allow us to calculate SASAs for protein structures and approximate how the SASA changes as the atomic coordinates move. These derivatives map each point in $\R^m$ to a corresponding gradient vector that we call \textbf{u}.

\begin{figure}[h!]
\caption{$B_1$ is bound by a sphere. The radius of the solvent is the distance between the edge of the ball and the sphere.}
\centerline{\includegraphics[width=3in]{Figure8}}
\end{figure}

@@

A similar expression to $\sigma_{ij}$ is derived for $\beta_{ijk}$, using $B^l_{ijk}$, the portion of the line segment $B_{ijk}$ on a fourth sphere $S_l$. 
\begin{equation}
\beta_{ijk} = 1 - \frac{\sum_{l}length(B^l_{ijk})}{length(B_{ijk})}
\end{equation}
In the right side term of equation 9, the sum adds up the length of the line segment that comes in contact with $S_l$ and divides that by the entire length of $B_{ijk}$ (which is $2\varrho_{ijk}$), giving the fraction of $F$ that is unexposed. By subtracting this term from 1, the fraction of solvent exposed area for the three intersecting spheres is found. We can think of the sphere as representing how much of the solvent can interact with the atom. 
\begin{comment}
The state \textbf{z} of each ball defines a center and a radius. As stated before, this paper considers that the weighted (E) and unweighted (A) areas are transformations from the space containing z ($\R^{3n}$) to $\R$. Their derivatives at a particular \textbf{z} are linear maps. In other words, the areas are represented by the map f and their derivatives the derivative of f at \textbf{z}.
\end{comment}

@@

These are the derivative equations that can then be tested on different molecules. It is important to note that there are cases where these equations will be discontinuous. We won't elaborate on these, but, for example, discontinuities can happen in a case where three spheres intersect at a circle. There are other cases of discontinuities, which will be useful in order successfully to run this algorithm for a variety of molecules.

@@

\section{Supplemental Information}
List of PDB's used: \\
129L, 1A70, 1A7H, 1AHL, 1AIU, 1AVS, 1B3C, 1B4R, 1B7I, 1B7K,\\
1B7V, 1B8W, 1BB8, 1BCG, 1BF0, 1BMG, 1BWR, 1CDT, 1COU, 1CPZ,\\
1CRE, 1DC9, 1DZO, 1EMV, 1FGP, 1FI5, 1FRE, 1G47, 1GAJ, 1HA6,\\
1HA8, 1HYM, 1I2U, 1I9Q, 1IIB, 1IVA, 1KTH, 1L6H, 1NEI, 1P3B,\\
1QYS, 1RKI, 1RO3, 1S2B, 1SRR, 1THX, 1UBQ, 1VHI, 1W4H, 1WDV,\\
1WOU, 2ACY, 2APB, 2BCB, 2CA8, 2CBM, 2CM8, 2EBY, 2FD7, 2FLS,\\
2GNP, 2GQV, 2GZV, 2HSH, 2KXK, 2MCT, 2O35, 2OXZ, 2R32, 2VV9,\\
2W5L, 2XKH, 2YM4, 2Z9H, 2ZO1, 3CS1, 3PFC, 3PMD, 3RQ8, 3S9B,\\
3UA6, 4AEI, 4BWS, 4DZN, 4FH1, 4KXO, 4LUP, 4W9A, 4ZK9, 5OH3,\\
5OIL, 5PC4, 5R2D, 6CQS, 6CW0, 6D40, 6DJA, 6F65, 6G0M, 6MKL.

>>>>

Our protein can be represented as a set of atom centers $S$ in $\R^3$ with their corresponding balls of various radii. One aspect of the model is computing the power diagram of this protein. The power diagram is a generalized Voronoi diagram, where instead of the normal Euclidean metric, the definition for distance between an atom with center $z_i$ and radius $\varrho_i$ and a point $x$ is $\pi_i(x)=\|x-z_i\|^2-\varrho_i^2$. This becomes the default Voronoi diagram of $S$ if all the radii are equal. After this power diagram is generated we can associate each set of atoms $S_i$ with its corresponding power cell $P_i$ defined by
\begin{equation*}
P_i=\{x\in\R^3\mid\pi_i(x)\leq\pi_j(x),\forall j\}
\end{equation*}
Each $P_i$ is convex, and together they cover all $\R^3$ without overlap. The dual of this power diagram, or alpha shape, is a weighted Delaunay Triangulation of $S$ into a set of simplices, or triangles and tetrahedrons. We can compute the surface area of these balls using Delaunay triangulation, which will help us in the later sections. There are additional rules for constructing Delaunay triangles, but for the purposes of this paper, it is only necessary to know how they are constructed from the atoms.

@@

Consider a set of balls $B_i$ defined by a center $z_i$ and a fixed radius $\varrho_i$ to be our protein. To represent the entire state of the protein (with $n$ atoms) we use the $m=3n$ dimensional  vector $\textbf{z}\in\R^{3n}$ where $\lbrack{\bf z}_{3i+1}, {\bf z}_{3i+2}, {\bf z}_{3i+3}\rbrack^T=z_i$. Note that $0\le i \le n-1$. Now we can define the weighted and unweighted surface areas $E,A:\R^{3m}\rightarrow\R$ as real multivariate functions from the state of the protein to a single value (the solvent exposed surface area). The weighted area will be described below. Here we will describe the gradients of these maps, written as $DE_{\bf z}$ and $DA_{\bf z}$ which would allow us to calculate SASAs for protein structures and approximate how the SASA changes as the atomic coordinates move. These derivatives map each point in $\R^m$ to a corresponding gradient vector that we call \textbf{u}.

\begin{figure}[h!]
\caption{$B_1$ is bound by a sphere. We can think of the sphere as how much of the solvent can interact with the atom.}
\centerline{\includegraphics[width=3in]{Figure8}}
\end{figure}

@@

A similar expression to $\sigma_{ij}$ is derived for $\beta_{ijk}$, using $B^l_{ijk}$, the portion of the line segment $B_{ijk}$ on a fourth sphere $S_l$. 
\begin{equation}
\beta_{ijk} = 1 - \frac{\sum_{l}length(B^l_{ijk})}{length(B_{ijk})}
\end{equation}
In the right side term of equation 9, the sum adds up the length of the line segment that comes in contact with $S_l$ and divides that by the entire length of $B_{ijk}$ (which is $2\varrho_{ijk}$), giving the fraction of $F$ that is unexposed. By subtracting this term from 1, the fraction of solvent exposed area for the three intersecting spheres is found.
\begin{comment}
The state \textbf{z} of each ball defines a center and a radius. As stated before, this paper considers that the weighted (E) and unweighted (A) areas are transformations from the space containing z ($\R^{3n}$) to $\R$. Their derivatives at a particular \textbf{z} are linear maps. In other words, the areas are represented by the map f and their derivatives the derivative of f at \textbf{z}.
\end{comment}

@@

These are the derivative equations that can then be tested on different molecules. It is important to note that there are cases where these equations will be discontinuous. We won't elaborate on these, but, for example, discontinuities can happen in a case where three spheres intersect at a circle. There are other cases of discontinuities, which will be useful to know in order to successfully run this algorithm for a variety of molecules.

@@

\section{Supplemental Information}
List of PDB's used: \\
1W4H, 2FD7, 1I2U, 1HA8, 1RO3, 1FRE, 1AHL, 1B8W, 1B7I, 3UA6,\\
1B7K, 1IVA, 1L6H, 1KTH, 2GQV, 2KXK, 1CDT, 1BF0, 1B7V, 4AEI,\\
4BWS, 1UBQ, 1FGP, 1CPZ, 1B4R, 1CRE, 2BCB, 4ZK9, 1BCG, 1B3C,\\
1G47, 1NEI, 1HYM, 1A7H, 1COU, 4W9A, 1A70, 1EMV, 1AVS, 1QYS,\\
1BB8, 1HA6, 2HSH, 2ACY, 1THX, 2Z9H, 2CBM, 2APB, 2FLS, 1AIU,\\
4DZN, 1SRR, 2GZV, 2XKH, 1RKI, 1BMG, 2MCT, 1FI5, 1WOU, 1DZO,\\
1DC9, 5PC4, 1WDV, 2OXZ, 1P3B, 129L, 4FH1, 1S2B, 2R32, 1BWR,\\
6MKL, 3CS1, 3PMD, 2O35, 3S9B, 3PFC, 2CA8, 2EBY, 1IIB, 1GAJ,\\
1I9Q, 4LUP, 3RQ8, 1VHI, 5OIL, 6CW0, 2YM4, 2ZO1, 4KXO, 6G0M,\\
5R2D, 2GNP, 6F65, 2CM8, 5OH3, 2W5L, 2VV9, 6CQS, 6DJA, 6D40.
